Space is the Place: Effects of Continuous Spatial Structure on Analysis of Population Genetic Data

### 2019-09-19 ms/spaceness.tex

We ran 400 simulations for the spatial and random-mating models on a square landscape of width $W=50$  \revpoint{AE}{3}
with per-unit carrying capacity $K=5$ (census $N \approx 10,000$), 
average lifetime $L=4$, 
genome size = $10^{8}$, 
recombination rate = $10^{-9}$, 
and drawing $\sigma$ values from a uniform distribution between 0.2 and 4. 
To speed up the simulations and limit memory overhead we used a mutation rate of 0 in SLiM 
and later applied mutations to the tree sequence with msprime's \texttt{mutate} function \citep{Kelleher2016}. 
Because msprime applies mutations proportionally to elapsed time,
we divided the mutation rate of $10^{-8}$ mutations per site per generation
by the average generation time estimated for each value of $\sigma$ (see `Demographic Parameters' below) to convert the rate to units of mutations per site per unit time.
(We verified that this procedure produced the correct number of mutations by comparing to a subset of simulations with SLiM-generated mutations, which are applied only at meiosis.)
Simulations were run for 1.6 million timesteps (approximately $30N$ generations).

@@

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% Supplement %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Supplementary Figures and Tables}
\beginsupplement

>>>>

We ran 400 simulations for the spatial and random-mating models on a square landscape of width $W=50$  \revpoint{AE}{3}
with per-unit carrying capacity $K=5$ (census $N \approx 10,000$), 
average lifetime $L=4$, 
genome size = $10^{8}$, 
recombination rate = $10^{-9}$, 
and drawing $\sigma$ values from a uniform distribution between 0.2 and 4. 
To speed up the simulations and limit memory overhead we used a mutation rate of 0 in SLiM 
and later applied mutations to the tree sequence with msprime's \texttt{mutate} function \citep{Kelleher2016}. 
Because msprime applies mutations proportionally to elapsed time,
we divided the mutation rate of $10^{-8}$ mutations per site per generation
by the average generation time estimated for each value of $\sigma$ (see `Demographic Parameters' below) to convert the rate to units of mutations per site per unit time.
We show that this procedure produces the same site-frequency spectrum as applying mutations directly in SLiM in Figure XXXXX. Simulations were run for 1.6 million timesteps (approximately $30N$ generations).

@@

Last, we used a subset of summary statistics to compare how our continuous model performs relative to a reverse-time stepping stone model (Appendix 1).

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% Supplement %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\beginappendix
\section{Appendix 1}

%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Comparisons with Stepping-Stone Models}
We checked that our model produces reasonable results by comparing it to a reverse-time stepping-stone model implemented in msprime \citep{Kelleher2016}. In this class of models we imagine an $n \times n$ grid of populations exchanging migrants with neighboring populations at rate $m$. If these models are good approximations of the continuous case we expect that results will converge as $n \to \infty$, so we ran simulations while varying $n$ from 5 to 50 (Table \ref{table:grid_params}). To compare with continuous models we first distributed the the same "effective" number of individuals across the landscape as in our continuous-space simulations ($\approx 6100$, estimated from $\theta_{\pi}$ of random-mating continuous-space simulations). We then approximate the mean per-generation dispersal distance $\sigma$ given a total landscape width $W$ as the product of the probability of an individual being a migrant and the distance traveled by migrants: $\sigma \approx 4m(W/n)$. We ran 500 simulations per $n$ while sampling $\sigma$ from $U(0.2,4)$. We then randomly selected 60 diploid individuals from each simulation (approximating diploidy by combining pairs of chromosomes with contiguous indices within demes) and calculated a set of six summary statistics using the scripts described in the summary statistics portion of the main text.

\begin{table}[h]
\centering
\begin{tabular}{ccc}
     demes per side ($n$)&$N_e$ per deme  & samples per deme  \\
     \hline
     5 & 244 & 20 \\
     10 & 61 & 10 \\
     20 & 15.25 & 2 \\
     50 & 2.44 & 1 \\
\end{tabular}
\captionsetup{justification=centering}
\caption{stepping-stone simulation parameters}
\label{table:grid_params}
\end{table}

\begin{figure*}[h]
\centering
\includegraphics[width=\textwidth]{ms/figures/grid_ss_curves.pdf}
\caption{Summary statistics for 2-dimensional coalescent stepping-stone models with fixed total $N_{e}$ and varying numbers of demes per side. The black "infinite" points are from our forward-time continuous space model. Inter-deme migration rates are related to $\sigma$ as described above.}
\label{fig:grid_stats}
\end{figure*}

In general we find many of the qualitative trends are similar among continuous and stepping-stone models and that, in most cases, statistics from stepping-stone models approach the continuous model as the resolution of the grid increases. For example, $\theta_\pi$ is inflated at low neighborhood sizes (i.e. low $m$), and the extent of the inflation increases to approach the continuous case as the resolution of the landscape increases. Similar pattterns are observed for $F_{is}$ and observed heterozygosity. However, $\theta_{W}$ behaves differently, with increased grid resolution leading to lower values. This in turn drives an even more positive Tajima's $D$ in grid simulations at small neighborhood sizes.

\begin{figure*}[h]
\centering
\includegraphics[]{ms/figures/grid_thetaW_tajD_zoom.pdf}
\caption{Zoomed-in view of $\theta_{W}$ and Tajima's $D$ for stepping-stone models under varying neighborhood sizes.}
\label{fig:grid_stats_zoom}
\end{figure*}

These differences relative to our continuous model mainly reflect two shortcomings of the reverse-time stepping stone model. If we simulate a coarse grid with relatively large populations in each deme, we cannot accurately capture the dynamics of small neighborhood sizes because mating within each deme remains random regardless of the migration rate connecting demes. This likely explains the trends in $\pi$, observed heterozygosity, and $F_{is}$. However increasing the number of demes while holding the total number of individuals constant results in small within-deme populations for which even the minimum sample size of 1 approaches the local $N_e$ (Table \ref{table:grid_params}). This results in an excess of short terminal branches in the coalescent tree, which decreases the total branch length and leads to fewer segregating sites, deflated $\theta_W$, and inflated Tajima's $D$. Overall then our continuous model reproduces important features of spatial structure approximated by reverse-time stepping-stone models at moderate neighborhood sizes while more accurately reflecting the dynamics of populations with small neighborhood sizes.

\stopappendix
\clearpage

\section{Supplementary Figures and Tables}
\beginsupplement
